---
title: "Assignment 2: Coding Basics"
author: "Jinglin Zhang"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
sequ <- seq(1,100,4) # generate sequence

#2. 
mean(sequ) # calculate the mean of sequence
median(sequ) # calculate the median of sequence 

#3. 
judger <- function (x){       # create a function to compare the mean and median
  if (mean(x) > median(x)){
    TRUE                      # if mean is greater than median, return 'TRUE'
  }
  else {
    FALSE                     # otherwise return 'FALSE'
  }
}                             
judger(sequ) 

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#install.packages('randomNames')
library(randomNames)
a <- randomNames(20) # character

b <- rnorm(20,60,15)
b <- round(b) # numeric

test <- function(x){
  ifelse(x<50,FALSE,TRUE)
}
c <- test(b) # logical

score <- data.frame(a,b,c)
colnames(score) <- c('Name','Score','Result')
score
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: both data frame and matrix can store two-dimensinal data in R, but matrix can only store one data type, while data frame can store multiple data types.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
test <- function(x){
  ifelse(x<50,FALSE,TRUE)
}
c <- test(b)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked, using if and else will reture the error "Error in if (x < 50) { : the condition has length > 1".the error occures because the if() function can only check 1 element in a vector, but we attemped to check the whole 20 elements in the vector.
